---
title: "ENVS-193DS Final"
date: "2026-06-02"
author: "Luke Simon"
format: 
  pdf:
    toc: true
execute:
  message: false
  warning: false
---

[Link to repo](https://github.com/lukesimonn/ENVS-193DS_spring-2026_final)

```{r}
#| label: packages-data

# read in packages
library(tidyverse)
library(here)
library(DHARMa)
library(ggeffects)

# reading in data and creating new object
nest <- read_csv(
  here("data", "nest_data_final.csv"))
```


# Problem 1. Research writing

# Problem 2. Data Analysis

## a.

```{r}
#| label: cleaning-wrangling

# creating clean data frame
nests_clean <- nest |> 
  # selecting desired columns
  select(ant_species, height_cm, case_control) |> 
  # filtering for desired ant species
  filter(ant_species %in% c("RRB", "BBR", "AB")) |> 
  # creating new column for full scientific name
  mutate(sci_name = case_when(
      ant_species == "RRB" ~ "Crematogaster sjostedti",
      ant_species == "BBR" ~ "Crematogaster mimosae",
      ant_species == "AB"  ~ "Crematogaster nigriceps"),
      # reordering factors for ant species
    ant_species = factor(ant_species,
      levels = c("RRB", "BBR", "AB"))
    )

# displaying structure of data frame
str(nests_clean)

# slicing sample
slice_sample(nests_clean, n = 10)
```

## b.

For this data, a value of 1 in the case_control column indicates that a nest was observed in the acacia tree in question. A value of zero in that said column means that no bird nest was observed in the tree.

## c.

### Tree Height (height_cm)

Tree height in this data set is a continuous variable that is measured in centimeters. As tree height increases the probability of nest occurrence would likely increase. Taller trees are more structurally complex and therefore have more nesting opportunities, they may also provide better predator avoidance for birds (Introduction, paragraph 2; Methods, "Study site and species" subsection, paragraph 1).

### Acacia ant species (ant_species)

As ant species changes, the probability of nest occurance would likely differ. This is due to each ant species exhibiting a distinct aggression level and mutualistic relationship with the host Acacia tree which affects the level of predator protection a nest in said tree would receive. (Introduction, paragraph 3; Methods, "Nest box experiment" subsection, paragraph 1)

## d.

| Model number | Ant Species | Tree Height | Predictor list                               |  
|--------------|:-----------:|:-----------:|:--------------------------------------------:|  
| 1            |     X       |      X      | all predictors, no interaction (null model)  |
| 2            |     X       |      X      | all predictors, with interaction (full model)|    

## e.

```{r}
#| echo: false

# model 1: all predictors no interaction
model1 <-glm(case_control ~ height_cm + ant_species,
             data = nests_clean,
             family = "binomial")

# model 2: all predictors with interaction
model2 <-glm(case_control ~ height_cm * ant_species,
             data = nests_clean,
             family = "binomial")
```

## f.

```{r}

# running AIC test to determine best model
AIC(model1, model2)
```

The best model as determined by Akaike’s Information Criterion (AIC) was the model that predicted test_case (whether or not a sample tree had a nest) using the interaction of tree height (height_cm) and species of acacia ant (ant_species)

## g.

```{r}

# plotting simulated residuals from DHARMa package.
plot(
  simulateResiduals(model2)
)
```

## h.

```{r}

# creating prediction model for model2
preds <- ggpredict(model2,
                   # terms for model, all tree heights all species
                   terms = c("height_cm[all]", "ant_species")) |> 
  # changing grouping column's name
  rename(ant_species = group) |>
  arrange(x)

# setting custom species colors
sp_cols <- c(
  "RRB" = "tan",
  "BBR" = "orange",
  "AB" = "lightblue3"
)

# creating facet labels from nest_clean
sp_labels <- nests_clean |>
  # choosing correct columns
  distinct(ant_species, sci_name) |>
  deframe()

# base layer: ggplot
ggplot() +
  # first layer: model confidence ribbon
  geom_ribbon(data = preds, # prediction data frame
              aes(x = x,
                  ymin = conf.low, # min by low confidence
                  ymax = conf.high, # max by high confidence
                  fill = ant_species)) + # filling color by group
  # second layer: model probability line
  geom_line(data = preds, # prediction data frame
            aes(x = x,
                y = predicted),
            # coloring line black
            color = "black",
            # making line thicker
            linewidth = 1) +
  # third layer: jitter for underlying data
  geom_jitter(data = nests_clean, # nests_clean data frame
              aes(x = height_cm, # x-axis: tree height
                  y = case_control, # y-axis: nest occurance
                  color = ant_species), # coloring by species
              alpha = 0.35, # making points more opaque
              shape = 21, # changing shape
              height = 0) + # removing vertical jitter
  # seperating by species
  facet_wrap(~ ant_species,
             # labelling with sci_name
             labeller = labeller(ant_species = sp_labels)) +
  # setting colors by species
  scale_color_manual(values = sp_cols) +
  scale_fill_manual(values  = sp_cols) +
  # relabelling x and y-axes
  labs(
    x = "Tree height (cm)", 
    y = "Probability of nest occurrence"
  ) +
  # changing theme
  theme_bw()+
  # removing legend
  theme(legend.position = "none",
        # removing gridlines
        panel.grid = element_blank()) 
``` 

## i.

**Figure 1. Model predictions for bird nest occurrence in acacia trees based on height and ant species** Figure displays model predictions for probability of nest occurrence as tree height increases, separated by ant species living on the tree. Line represents exact predictions of model that assumes interaction between species and tree height. Ribbon represents model 95% confidence interval. Species are sorted by color with *C. sjostediti* being tan, *C. mimosae* being orange, and *C. nigriceps* being blue. Data originally from Lujan, E., Nielsen, R., Short, Z., Wicks, S., Nderitu Watetu, W., Malingati Khasoha, L., Palmer, T. M., Goheen, J. R., & Alston, J. M. (2023). Data, code, and metadata for: Symbiotic acacia ants drive nesting behavior by birds in an African savanna [Data set]. Zenodo. https://doi.org/10.5281/zenodo.8373322.

## j.

```{r}

# creating prediction at 600 cm
predict_600 <- ggpredict(model2,
                         terms = c("height_cm [600]", "ant_species"))
```

## k.


# Problem 3. Problem 3. Affective and exploratory visualizations

a.

